Structured Versus Decorated Cospans

### 2020-11-26 cospans.tex
Trying to help with Section 5.4

%%% Do reaction networks and dynamical systems (which do NOT come from structured cospans).

>>>>

%%% Do reaction networks and dynamical systems (which do NOT come from structured cospans).
A dynamical system is a vector field from the perspective of a system of first-order ordinary differential equations.
Chemical reaction networks are a specific kind of dynamical system in which the vector field is algebraic, meaning that its coefficients are polynomials.
Using Fong's original incarnation of decorated cospans, Pollard and the first author \cite{BP} constructed a symmetric monoidal category $\mathsf{OpenDynam}$ for which the morphisms are open dynamical systems. This category $\mathsf{OpenDynam}$ has:
\begin{itemize}
\item{objects as finite sets, and}
\item{morphisms as equivalence classes of cospans of finite sets whose apices are equipped with an algebraic vector field.
\[
\begin{tikzpicture}[scale=1.5]
\node (D) at (-3,0) {$a$};
\node (E) at (-2,0) {$s$};
\node (F) at (-1,0) {$b$};
\node (A) at (0,0) {$v_s \colon \lR^s \to \lR^s$};
\path[->,font=\scriptsize,>=angle 90]
(D) edge node [above] {$i$} (E)
(F) edge node [above] {$o$} (E);
\end{tikzpicture}
\]
}
\end{itemize}
This category is constructed from a symmetric lax monoidal functor $F \colon \Fin\Set \to \Set$ that assigns to a finite set $s$ the set $F(s)$ of algebraic vector fields on $s$ and a to function $f \colon s \to s'$ the function $F(f) \colon F(s) \to F(s')$ defined by $$F(f)(v_s)=f_* v_s f^*$$
where $f^\ast \colon \mathbb{R}^{s'} \to \mathbb{R}^{s}$ is the \define{pullback} defined by 
$$f^\ast (c_{s'})(\sigma)=c_{s}(f(\sigma))$$ 
and $f_{\ast} \colon \mathbb{R}^s \to \mathbb{R}^{s'}$ is the \define{pushforward} defined by 
$$f_{\ast}(c_{s'})(\sigma^\prime) = \sum_{\{\sigma \in s : \; f(\sigma) = \sigma^\prime\}} c_{s}(\sigma)$$
for $c_{s} \in \mathbb{R}^{s}$ and $c_{s'} \in \mathbb{R}^{s'}$
Functoriality of $F$ comes the pushforward being covariant and the pullback being contravariant, and the lax monoidal structure of $F$ stems from the canonical isomorphism $\lR^s \times \lR^{s'} \cong \lR^{s+s'}$ which is used to map a pair of vector fields $v_s \colon \lR^s \to \lR^{s}$ and $v_{s'} \colon \lR^{s'} \to \lR^{s'}$ to the vector field $v_{s+s'} \colon \lR^{s+s'} \to \lR^{s+s'}$. We now show that open dynamical systems of this nature are an example of open systems that are able to be characterized by decorated cospans but not structured cospans.

Let $\mathsf{Dynam}$ be the category for which:
\begin{itemize}
\item{an object is a pair $(s,v_s)$ where $s$ is a finite set and $v_s$ is an algebraic vector field $v_s \colon \lR^s \to \lR^s$ on the finite set $s$, and}
\item{a morphism from $(s,v_s)$ to $(s',v_{s'})$ is a function $f \colon s \to s'$ such that $v_{s'} = f_* v_s f^*$.}
\end{itemize}

A well-known result is the following:
\begin{lem}
A functor $U \colon \A \to \X$ admits a left adjoint if and only if for every $x \in \X$, the comma category $x \downarrow U$ has an initial object.
\end{lem}
Using this result, one can show that the forgetful functor $U \colon \mathsf{Dynam} \to \Fin\Set$ is not a right adjoint. In particular, by taking $x$ to be the empty set, the comma category $\emptyset \downarrow U$ has:
\begin{itemize}
\item{objects given as pairs $((s,v_s),! \colon \emptyset \to s)$, and}
\item{a morphism from $((s,v_s),! \colon \emptyset \to s)$ to $((s',v_{s'}),! \colon \emptyset \to s')$ is a morphism $f \colon (s,v_s) \to (s',v_{s'})$ in $\mathsf{Dynam}$ such that $U(f) \circ ! = !$ in $\Fin\Set$, a condition which amounts to simply being a morphism in $\mathsf{Dynam}$.}
\end{itemize}
In particular, the object $((\emptyset,v_\emptyset),! \colon \emptyset \to \emptyset)$ resides in the comma category $\emptyset \downarrow U$, where $v_\emptyset$ is the only possible vector field on the empty set --- namely, the dengerate vector field. Given any other object $((s,v_s),! \colon \emptyset \to s)$ in $\emptyset \downarrow U$, there exists no morphisms $! \colon \emptyset \to s$ from the first object to the second unless $s$ is also the empty set as the vector field $$!_* v_\emptyset !^*$$will also always be degenerate and not equal to $v_s$. Clearly there are also no morphisms in the other direction for $s \neq \emptyset$ as $f$ must be a function. Thus the forgetful functor $U$ has no left adjoint $L \colon \Fin\Set \to \mathsf{Dynam}$, and hence we cannot construct a structured cospan category ${_L \Csp(\mathsf{Dynam})}$ whose morphisms are equivalence classes of open dynamical systems of this type.
